Competing Constructions for German Measure Phrases

### 2017-03-16 Latex/article.tex
work on previous analyses

This paper deals with a morpho-syntactic alternation that occurs only in a very specific syntactic measure noun phrase construction.
By \textit{alternation} I refer to a situation where two or more forms or constructions are available with no clear difference in acceptability, function, or meaning.%
\footnote{In German grammar, the term \textit{Zweifelsfälle} `cases of doubt' is often used \citep{Klein2009}.}
I use it as an example case to introduce a protocol for the estimation and validation of corpus-based models in cognitively oriented corpus linguistics.

@@

\subsection{Two stable cases and a case alternation}

The alternation phenomenon to be discussed in this paper appears to be a simple case alternation.
However, without a solid knowledge of German nominal morpho-syntax, some of the more subtle distinctions to be made cudl confuse readers.
Therefore, I introduce and illustrate the relevant constructions rather thoroughly in this section, including a review of the relevant literature.
The discussion defines the narrow range of syntactic constellations in which the alternation occurs and thus motivates the focus on \textit{only} this narrow range rather than, for example, the whole range of nominal constructions expressing quantities.

I use the term \textit{measure noun phrase} (MNP) to refer to noun phrases (NPs) where a kind-denoting (count or mass) noun depends on another noun specifying a quantity of the objects or the substance denoted by the kind noun.
I call the kind-denoting noun simply the \textit{kind noun} and the quantity-denoting noun the \textit{measure noun}.%
\footnote{It would be more technically appropriate -- but also more cumbersome -- to speak of the \textit{kind noun phrase} because the embedded noun can be modified, cf.\ below.}
Measure nouns can be all sorts of nouns which genuinely denote a quantity (such as \textit{litre} or \textit{amount}) but also nouns denoting containers, collections, etc. (such as \textit{glass} or \textit{bucket}).
In a similar vein, \cite[284]{Brems2003} also considers nouns as measure nouns \textit{which, strictly speaking, do not designate a `measure', but display a more nebulous} (\textit{sic!}) \textit{potential for quantification}.

Descriptively, three different situations need to be distinguished \wrt case assignment in German measure noun phrases.
If the kind noun comes with a determiner, the construction resembles (and is usually called) a \textit{pseudo-partitive} (on partitives and pseudo-partitives, e.\,g., \citealp{Barker1998,Selkirk1977,Stickney2007,Vos1999}; for a recent application of the terminology to German, see \citealp{Gerstenberger2015}).%
\footnote{Whereas partitives are constructions denoting a proper part-of relation such as \textit{a sip of the wine}, pseudo partitives -- albeit syntactically similar and diachronically related to partitives in many languages -- merely denote quantities such as \textit{a sip of wine}.
See the references just given.
In the literature on German, some authors incorrectly call the pseudo-partitive a \textit{partitive} \citep{Zimmer2015} while some realise the difference \citep[225]{Eschenbach1994}.}
I use * to denote general unacceptability, and the acceptability judgements given in (\ref{ex:intro:pseudopartitive}) are completely undisputed.
Notice that the kind noun is obligatory in the genitive and that case agreement with the measure noun is not an option (but see immediately below).
I refer to the construction in (\ref{ex:intro:pseudopartitive1}) as the \textit{Pseudo-partitive Genitive Construction} (PGC).

@@

When the kind noun is bare -- \ie\ when it comes without a determiner and without modifying adjectives -- the acceptability of the case patterns is reversed, and the only acceptable construction is usually classified as a \textit{Narrow Apposition Construction} \citep{Loebel1986}, henceforth NAC.
In the NAC, the kind noun agrees in case with the measure noun, see (\ref{ex:intro:narrowapposition}).
Notice that the unavailability of the genitive on the kind noun follows from a quirky general constraint that genitive NPs in German require the presence of some strongly case-marked element (determiner or adjective) in addition to the head noun in order to be acceptable \citep{GallmannLindauer1994,Schachtl1989}.

@@

The construction as in (\ref{ex:intro:narrowapposition2}) is also referred to as the \textit{Direct Partitive Construction} for other Germanic languages where the PGC with the synthetic genitive is not available.%
\footnote{This nomenclature makes sense in contrast to the \textit{Indirect Partitive Construction} with prepositional linkers translating to `of' -- \ie\ analytic genitives -- in such languages, see \cite{HankamerMikkelsen2008} for Danish.
For German, this terminology is not distinctive enough, which is why I use the terms NAC and PGC.}
While only the accusative is shown, case agreement of the kind noun is obligatory also with nominative and dative measure nouns.

The actual alternation can be observed \textit{only when the kind noun occurs with an attributive adjective but without a determiner}, as in (\ref{ex:intro:alternation}), where both the NAC in (\ref{ex:intro:alternation1}) and the PGC in (\ref{ex:intro:alternation2}) are not only both acceptable but also functionally and semantically identical.%
\footnote{Some more normative grammars take stronger positions w.\,r.\,t. the grammaticality of these two options.
  See \cite{Hentschel1993,Zimmer2015} for analyses of the stances taken in more or less descriptive or normative grammars of German.
  As the usage and experimental data presented below should render it clear, there might be preferences under certain circumstances, but we cannot assume either construction to be \textit{ungrammatical} or -- for readers who reject the notion of grammaticality: \textit{(virtually) unanimously unacceptable}.}
The distribution of the case patterns of NAC (case identity between measure noun and kind noun) an PGC (genitive on kind noun) depending on the structure of the kind NP are summarised in Table~\ref{tab:constructions}.

@@

\begin{table}
  \centering
  \begin{tabular}{lccc}
    \multicolumn{1}{r}{kind NP:} & bare noun & with adjective & with determiner \\
    \midrule
    narrow apposition         & \CheckIt & \CheckIt &          \\
    pseudo-partitive genitive &          & \CheckIt & \CheckIt \\
  \end{tabular}
  \caption{Distribution of the NAC and PGC constructions in different NP structures}
  \label{tab:constructions}
\end{table}

I now turn to some subtler points related to the measure noun case alternation, namely:

\begin{enumerate}[i.]
  \item{\label{it:intro:datsg} alternative forms of dative singular neuter adjectives}
  \item{\label{it:intro:femsg} case syncretism in feminine NPs}
  \item{\label{it:intro:plurl} the parallel plural/collective construction(s)}
  \item{\label{it:intro:noifl} grammaticalised non-inflected measure nouns}
  \item{\label{it:intro:other} alternative constructions for expressing quantities}
\end{enumerate}

As for \ref{it:intro:datsg}, a complication with neuter kind nouns in the dative is mentioned by \cite[20--22]{Zimmer2015}.
He reports a high number of occurrences of adjectives being inflected according to the \textit{weak} inflectional pattern, which is normally only used if a strongly inflected determiner precedes the adjective: \textit{rot-en Wein} instead of \textit{rot-em Wein}.
In my corpus sample (to be made available after publication of this paper), this tendency was not nearly as clear, and there was a high number of very noisy sentences among those potentially showing this pattern.
Consequently, I do not discuss the problem further.

With feminine kind nouns, \ref{it:intro:femsg} comes into play.
Whereas the singular masculine and neuter nominal subsystem still represents a four case system with a distinguishable genitive, feminine singular NPs show nominative-accusative and dative-genitive syncretisms.
Table~\ref{tab:syncretisms} also shows the partial syncretism in the plural, where no gender distinctions are made (for an overview, see \citealp[XYZ]{Eisenberg2013a}).
This means that with dative measure nouns, the alternation is unobservable if the kind noun is feminine, and it is problematic to speak of a \textit{genitive} kind noun construction in this case.
This will be taken into account in Section~\ref{sec:corpusmodels} when the corpus data are discussed.

@@

Turning to \ref{it:intro:plurl}, readers might habe notices that so far only MNPs denoting quantities of substances (mass kind nouns such as \textit{ein Glas roter Wein\slash roten Weines} `a glass of red wine') have been discussed.
If the kind noun is a plural count noun with a collective denotation such as \textit{ein Sack kleine Äpfel\slash kleiner Äpfel} `a bag of small apples' (see Table~\ref{tab:syncretisms} for the inflection), a similar alternation between PGC and NAC can be observed.%
\footnote{On the collective interpretation of such constructions, see \cite{Eschenbach1994}.
It is, however, very surprising that she does not notice the morpho-syntactic distribution of PGC and NAC, even when she gets very close to listing the relevant data on p.\ 217.}
In line with experimental results reported in \cite[15--16]{Zimmer2015}, I found that the PGC is so dominant with plural kind nouns (over 92\%, cf.\ Section~\ref{sec:corpusmodels}) that the alternation is virtually unanalysable.
While this will play a role in the interpretation of the corpus findings, such plural\slash collective MNPs will not be dealt with prominently in the remainder of the paper.

Turning to \ref{it:intro:noifl}, some measure nouns have been grammaticalised in a way that they always appear non-inflected.
They are typical measure nouns like \textit{Gramm} `gram', \textit{Pfund} `pound' or \textit{Prozent} `percent'.%
\footnote{With neglectable frequency, this effect is extended to container nouns like \textit{Glas} `glass'.
A non-inflected variant \textit{zwei Glas roter Wein} and an inflected variant \textit{zwei Gläser roter Wein} `two glasses of wine' co-exist.
The non-inflected variant denotes the quantity of wine fitting into two glasses, the inflected variant denotes two glasses filled with wine.}
I treat these cases like any other measure noun because they enter into both the NAC as in (\ref{ex:uninflectedmeasures:a}) and the PGC as in (\ref{ex:uninflectedmeasures:b}).
The statistical models presented in Section~\ref{sec:corpusmodels} will, however, show whether they (or other highly grammaticalised measure nouns) favor the NAC or the PGC.

@@

Finally, \ref{it:intro:other} suggests that there are alternative ways of expressing similar quantifications.
First, the analytic (pseudo-)partitive with \textit{von} `of' is only available as an alternative to the PGC if the kind noun phrase contains a determiner, cf. (\ref{ex:analyticalpartitive}).

@@

However, if the measure noun denotes a container, a construction with \textit{voll} (\textit{von})\slash\textit{voller} `full (of)' is available, see (\ref{ex:voll}).
This construction has very specific properties discussed in \cite{Zeldes2018}.
While subtle interactions between the \textit{voll}\slash\textit{voller} constructions and NAC\slash PGC cannot be excluded, there are no clear and motivated hypotheses about such interactions, and I do not discuss the construction further.

@@

This concludes the descriptive overview of the alternation.

\subsection{Analyses of the constructions}

As stated in Section~\ref{sec:cogocl}, I consider it essential not to disregard generalisations discovered in descriptive, generative or formal frameworks, even if the accompanying formulations do not fit into probabilistic settings.
This section therefore reviews some existing analyses of the PGC-NAC alternation or related issues, including generative approaches.

% Grammaticalisation (Brems, Rutkowski)
% Syntactic structure: headedness (Galmann>Löbel, Eschenbach p. 213, Vos: DPC/DCC), difference to English (Brems)
% Syntactic structure: adjectives-as-determiners and layers in NP, role assignment to nouns in PGC/NAC
%     Gallmann>Bhatt1990 DP[gen] vs NP[agree]
% Influencing factors: Hentschel, Zimmer
% Squeeze in Anttila & Fong

@@

@@

\subsection{Statistical issues}
\label{ssec:issuesstatistics}

@@

In the following, I examine the case alternation in German measure NPs under a cognitively oriented corpus linguistic perspective.
Since, however, the statistical tool kit (hierarchical generalised linear models, general tools such as $p$ values and confidence intervals, resampling techniques, model selection and model evaluation strategies, alternative algorithms for estimation, etc.) is heterogeneous and often criticised (especially under noisy conditions and when researcher run the risk of doing data mining instead of proper theory testing) I present a careful protocol for specifying, estimating, validating, and interpreting such models in Section \ref{sec:corpusmodels}.
Before working out the models, I show how I made sure that the relevant data is contained in the sample in Section \ref{sec:gettingdata}.
Furthermore, to provide external validation, I present two experiments in Section \ref{sec:externalvalidation} in which I attempted to find correlations between the corpus-based findings and speakers' behaviour.

@@

\subsection{Tools for modelling alternations}
\label{sec:righttools}

In this section, I discuss the relevant statistical approaches to modelling alternations based on corpus data.
I mainly focus on well-established Generalised Linear Models (GLMs) and their \textit{multi-level} or \textit{hierarchical} extensions, which are also called \textit{mixed models} (GLMMs).

% TODO CONTENT HERE

In Section~\ref{ssec:corpusnonhierarchicalmodel}, I briefly discuss why a non-hierarchical model -- while it is an option in principle -- has serious disadvantages for the data set under discussion (and similar data sets).
In Section~\ref{ssec:corpushierarchicalmodel}, I present the hierarchical model which is actually interpreted as the primary result of this study.
Section~\ref{ssec:bayesian} takes a short detour and shows (by example) that results from Bayesian estimation are not intrinsically different from those obtained from Maximum Likelihood methods, although Bayesian approaches have recently been touted to solve various problems facing applied statistics in various fields, even including cognitively oriented corpus linguistics.
Finally, I summarise and interpret the results of the corpus study in Section~\ref{ssec:modelssummary}.

>>>>

This paper deals with a morpho-syntactic alternation that occurs only in a very specific syntactic measure noun phrase construction.
By \textit{alternation} I refer to a situation where two or more forms or constructions are available with no clear difference in acceptability, function, or meaning.%
\footnote{In German grammar, the term \textit{Zweifelsfälle} `cases of doubt' is often used \citep{Klein2009}.}
I use it as an example case to introduce a protocol for the estimation and validation of corpus-based models in cognitively oriented corpus linguistics which I briefly motivate and describe in this introduction.

% Add to CogL special issue: Look at what non-cognitivists do
% => their have interesting data and interesting generalisations (which are not intrinsically nonsense or incompatible with CogLi)
% => where they lack is modelling probabilistic effects, and that should be one of the strengths of cognitively oriented (corpus) linguistics
% => FUZZINESS and SIMILARITY in classification (prototypes and exemplars)
% => advantage: testable with corpus and experiment (ideally both)

% This is a case study and more.
% => gathering a descriptive picture from non-cognitive literature
% => motivating the phenomenon as relevant for CogLi
% => careful protocol for:
%  =>> motivating range of alternants under examination and selecting hypotheses (sound and/or slightly less sound, but MAKE CLEAR what they are!)
%  =>> motivate whether larger utterance context plays a role or not (sociolinguistic & register(?) commitment)
%  =>> choosing right data
%  =>> choosing "the right" – NO, "some appropriate" statistics
%  =>> getting minimal external validation with experiments

% This article attempts to follow this protocol as well as possible given the available analyses/data sources/nature of the phenomenon.

@@

\subsection{Two stable cases and a case alternation}
\label{sec:descriptive}

The alternation phenomenon to be discussed in this paper appears to be a simple case alternation.
However, without a solid knowledge of German nominal morpho-syntax, some of the more subtle distinctions to be made can be confusing.
Therefore, I introduce and illustrate the relevant constructions in some detail in this section, including a review of the relevant literature.
The discussion defines the narrow range of syntactic constellations in which the alternation occurs and thus motivates the focus on \textit{only} this narrow range rather than the whole range of nominal constructions expressing quantities.

I use the term \textit{measure noun phrase} (MNP) to refer a noun phrase (NP) in which a kind-denoting (count or mass) noun depends on another noun specifying a quantity of the objects or the substance denoted by the kind noun.
I call the kind-denoting noun simply the \textit{kind noun} and the quantity-denoting noun the \textit{measure noun}.%
\footnote{It would be more technically appropriate -- but also more cumbersome -- to speak of the \textit{kind noun phrase} because the embedded noun can be modified and form a full noun phrase, cf.\ immediately below and Section~\ref{sec:analyses} on the phrasal status of the constituents.}
Measure nouns can be all sorts of nouns which genuinely denote a quantity (such as \textit{litre} or \textit{amount}) but also nouns denoting containers, collections, etc. (such as \textit{glass} or \textit{bucket}).
In a similar vein, \cite[284]{Brems2003} considers nouns as measure nouns \textit{which, strictly speaking, do not designate a `measure', but display a more nebulous} (\textit{sic!}) \textit{potential for quantification} (see also \citealp[530]{Koptjevskaja2001}, and \citealp[338]{Rutkowski2007}).
For illustration purposes, in the English \textit{a cup of fine coffee}, \textit{cup} is the measure noun, and \textit{coffee} is the kind noun.

Descriptively, three different syntactic configurations need to be distinguished \wrt case assignment inside German measure noun phrases.
If the kind noun comes with a determiner, the construction resembles (and is usually called) a \textit{pseudo-partitive} (on partitives and pseudo-partitives see, e.\,g., \citealp{Barker1998,Selkirk1977,Stickney2007,Vos1999}; for a recent application of the terminology to German, see \citealp{Gerstenberger2015}).%
\footnote{Whereas partitives are constructions denoting a proper part-of relation such as \textit{a sip of the wine}, pseudo partitives -- albeit syntactically similar and diachronically related to partitives in many languages -- merely denote quantities such as \textit{a sip of wine}.
In the literature on German, some authors incorrectly call the pseudo-partitive a \textit{partitive} \citep{Zimmer2015} while some realise the difference \citep[225]{Eschenbach1994}.}
I use * to denote general unacceptability, and the acceptability judgements given in (\ref{ex:intro:pseudopartitive}) are completely undisputed.
Notice that the kind noun is obligatory in the genitive and that case agreement with the measure noun is not an option (but see immediately below).
I refer to the construction in (\ref{ex:intro:pseudopartitive1}) as the \textit{Pseudo-partitive Genitive Construction} (PGC).

@@

When the kind noun is bare -- \ie\ when it comes without a determiner and without modifying adjectives -- the acceptability of the case patterns is reversed, and the only acceptable construction is usually classified as a \textit{Narrow Apposition Construction} \citep{Loebel1986}, henceforth NAC.
In the NAC, the kind noun agrees in case with the measure noun, see (\ref{ex:intro:narrowapposition}).
Notice that the unavailability of the genitive on the kind noun follows from a quirky constraint that genitive NPs in German require the presence of some strongly case-marked element (determiner or adjective) in addition to the head noun in order to be acceptable \citep{GallmannLindauer1994,Schachtl1989}.

@@

The construction as in (\ref{ex:intro:narrowapposition2}) is also referred to as the \textit{Direct Partitive Construction} for other Germanic languages where the PGC with the synthetic genitive is not available.%
\footnote{This nomenclature makes sense in contrast to the \textit{Indirect Partitive Construction} with prepositional linkers translating to `of' -- \ie\ analytic genitives -- in such languages, see \cite{HankamerMikkelsen2008} for Danish.
For German, this terminology is not distinctive enough, which is why I use the terms NAC and PGC.}
While only the accusative is shown, the kind noun obligatory agrees in case also with nominative and dative measure nouns.

The actual alternation can be observed \textit{only when the kind noun occurs with an attributive adjective but without a determiner}, as in (\ref{ex:intro:alternation}), where both the NAC in (\ref{ex:intro:alternation1}) and the PGC in (\ref{ex:intro:alternation2}) are equally acceptable.
They are also functionally and semantically equivalent.%
\footnote{Some descriptive and normative grammars take stronger positions w.\,r.\,t. the grammaticality of these two options.
See \cite{Hentschel1993,Zimmer2015} for analyses of the sometimes absurd stances taken in grammars of German.
As the usage and experimental data presented below should render it clear, there might be preferences under certain circumstances, but we cannot assume either construction to be unacceptable.}

@@

\begin{table}
  \centering
  \begin{tabular}{lccc}
    \multicolumn{1}{r}{kind NP:} & bare noun & with adjective & with determiner \\
    & [\ldots N\Subsf{meas} [N\Subsf{kind}]] & [\ldots N\Subsf{meas} [AP N\Subsf{kind}]] & [\ldots N\Subsf{meas} [D N\Subsf{kind}]] \\
    \midrule
    narrow apposition         & \CheckIt & \CheckIt &          \\
    pseudo-partitive genitive &          & \CheckIt & \CheckIt \\
  \end{tabular}
  \caption{Distribution of the NAC and PGC constructions in different NP structures}
  \label{tab:constructions}
\end{table}

The distribution of the case patterns in the NAC (case identity between the measure noun and the kind noun) and in the PGC (genitive on the kind noun) depending on the structure of the kind NP are summarised in Table~\ref{tab:constructions}.
I now turn to some more subtle issue related to the measure noun case alternation, namely:

\begin{enumerate}[i.]
  \item{\label{it:intro:datsg} alternative \textit{weak} forms of dative singular neuter adjectives}
  \item{\label{it:intro:femsg} case syncretism in feminine NPs}
  \item{\label{it:intro:plurl} similar plural/collective construction(s)}
  \item{\label{it:intro:noifl} grammaticalised non-inflected measure nouns}
  \item{\label{it:intro:other} alternative constructions for expressing quantities}
\end{enumerate}

As for issue \ref{it:intro:datsg}, a complication with neuter kind nouns in the dative is mentioned by \cite[20--22]{Zimmer2015}.
He reports a high number of occurrences of adjectives being inflected according to the \textit{weak} inflectional pattern, which is normally only used if a strongly inflected determiner precedes the adjective: \textit{kalt-en Wasser} `cold water' instead of \textit{kalt-em Wasser}.%
\footnote{The \textit{strong} case and number suffixes of the adjectives are (with one exception) the suffixes of articles and pronouns, and they are used when no strongly inflected pronoun or determiner precedes the adjective.}
In my corpus sample (released with this article), this tendency was not nearly as clear, and there was a high number of very noisy sentences among those potentially showing this pattern.
Also, in my syntactic\slash semantic interpretation of the alternation (see Section~\ref{sec:corpusmodels}), the alternative weak forms are not expected to influence the alternation.
Consequently, I do not discuss the problem further.

\label{page:femininesyncretism}
With feminine kind nouns, issue \ref{it:intro:femsg} becomes relevant.
Whereas the singular masculine and neuter nominal subsystem is a four case system (at least for NPs with adjectives and determiners), feminine singular NPs show nominative-accusative and dative-genitive syncretisms.
Table~\ref{tab:syncretisms} shows this and the partial syncretism in the plural, where no gender distinctions are made (for an overview, see \citealp[XYZ]{Eisenberg2013a}).
This means that with dative measure nouns, the alternation is unobservable if the kind noun is feminine, and it is problematic to speak of a \textit{genitive} kind noun construction in this case.
This will be taken into account in Section~\ref{sec:corpusmodels} when the corpus data are discussed.

@@

Turning to \ref{it:intro:plurl}, readers might have notices that so far only MNPs denoting quantities of substances (mass kind nouns such as \textit{ein Glas roter Wein\slash roten Weines} `a glass of red wine') have been discussed.
If the kind noun is a plural count noun with a collective denotation such as \textit{ein Sack kleine Äpfel\slash kleiner Äpfel} `a bag of small apples' (see Table~\ref{tab:syncretisms} for the inflection), a similar alternation between PGC and NAC can be observed.%
\footnote{\label{fn:eschenbash}On the collective interpretation of such constructions, see \cite{Eschenbach1994}.
It is, however, very surprising that Eschenbach does not notice and discuss the morpho-syntactic distribution of PGC and NAC, even when she gets very close to listing the relevant data \citep[217]{Eschenbach1994}.}
In line with experimental results reported in \cite[15--16]{Zimmer2015}, I found that the PGC is so dominant with plural kind nouns (over 92\%, cf.\ Section~\ref{sec:corpusmodels}) that the alternation cannot be analysed in the same way as in the singular case.
While this will play a role in the interpretation of the corpus findings, such plural\slash collective MNPs will not be dealt with prominently in the remainder of the paper.
A statistical model for plural kind nouns will be reported in passing in Section~\ref{sec:corpusmodels}, however.

Turning to \ref{it:intro:noifl}, some measure nouns have been grammaticalised in a way that they always appear non-inflected.
They are typical measure nouns like \textit{Gramm} `gram', \textit{Pfund} `pound' or \textit{Prozent} `percent'.%
\footnote{With neglectable frequency, this effect is extended to container nouns like \textit{Glas} `glass', but with a twist.
A non-inflected variant \textit{zwei Glas roter Wein} `two glasses of wine' and an inflected variant \textit{zwei Gläser roter Wein} co-exist.
The non-inflected variant denotes the quantity of wine fitting into two glasses, the inflected variant denotes two glasses filled with wine.}
I treat these cases like any other measure noun because they enter into both the NAC as in (\ref{ex:uninflectedmeasures:a}) and the PGC as in (\ref{ex:uninflectedmeasures:b}).
The statistical models presented in Section~\ref{sec:corpusmodels} will, however, show whether they (or other highly grammaticalised measure nouns) favor the NAC or the PGC through the estimation of lemma-specific random intercepts.

@@

Finally, \ref{it:intro:other} suggests that there are alternative ways of expressing quantification.
First, the analytic (pseudo-)partitive with \textit{von} `of' is only available as an alternative to the PGC if the kind noun phrase contains a (definite or indefinite) determiner as in (\ref{ex:analyticalpartitive}).

@@

Second, however, if the measure noun denotes a container, a construction with \textit{voll} (\textit{von})\slash\textit{voller} `full (of)' is available, see (\ref{ex:voll}).

@@

This construction has very specific properties discussed in \cite{Zeldes2018}, and obviously is .
While subtle interactions between the \textit{voll}\slash\textit{voller} constructions and NAC\slash PGC cannot be excluded, there are no clear and motivated hypotheses about such interactions, and I do not discuss the construction further.
This concludes the descriptive overview of the alternation.

\subsection{Previous analyses}
\label{sec:analyses}

This section reviews some existing analyses of the PGC-NAC alternation or related issues.
I do \textit{not} review any of the semantics literature on partitives vs.\ pseudo-partitives because the constructions of interest here are all semantically pseudo-partitives, and thus none of the variance in the choice of the PGC and NAC can be expected to be explained by a discussion along these lines.
The same is mostly true for the work on the grammaticalisation paths leading to the formation of pseudo-partitives (\eg\ \citealp{Brems2003,DeclerckBrems2016,Koptjevskaja2001,Rutkowski2007}).
It is often assumed that pseudo-partitives arise as a form of grammaticalised partitives (\eg\ \citealp[536--539]{Koptjevskaja2001} for Finnish and Estonian, \citealp[559]{Koptjevskaja2001} for European languages per se).%
\footnote{Notice that the description of German in \cite[547--549]{Koptjevskaja2001} (mostly referring \citealp{Eschenbach1993}) is partly incorrect, maybe because her main source \cite{Eschenbach1993} is not as clear as it could be (see also Footnote \ref{fn:eschenbash} above).
Her example (29b) from \cite[71]{Eschenbach1993}, repeated here verbatim as (\ref{ex:koptjewskaja}) is misclassified as a clear case of a dative on the measure noun and a genitive on the kind noun.

\begin{exe}
  \ex{\label{ex:koptjewskaja}\gll trotz einer Flasche gut-en Wein-s\\
  in.spite.of one-\textsc{fem.gen} bottle good-\textsc{masc.sg.gen} wine-\textsc{gen}\\
  \glt in spite of one bottle of good wine}
\end{exe}

It could, however, equally well be a genitive on the measure noun because of the syncretisms discussed here on p.\ \ref{page:femininesyncretism} and an ambiguity of the preposition \textit{troz}.
Then, she does not clearly distinguish acceptable from non-acceptable constructions (cf.\ Section \ref{sec:descriptive}) on her pp.\ 548 and 549, including a rather confusing \textit{???} rating of her (31b), which is clearly a non-acceptable construction of the type illustrated in (\ref{ex:analyticalpartitive:b}) above.
}

% Syntactic structure: headedness (Galmann>Löbel, Eschenbach p. 213, Vos: DPC/DCC), difference to English (Brems)
% Syntactic structure: adjectives-as-determiners and layers in NP, role assignment to nouns in PGC/NAC
%     Gallmann>Bhatt1990 DP[gen] vs NP[agree]
% Influencing factors: Hentschel, Zimmer
% Squeeze in Anttila & Fong

@@

% Final statement: This is about choosing case in a specific configuration, not about choosing one construction from a broader range of constructions expressing quantification/measure. Maybe mention variable/variant and suggest alternation vs. variation as terminology. Mention that there are NO specific hypotheses about socio-linguistic influences and maybe weaker style/register hypotheses MIXED with semantic hypotheses about classes of kind nouns (which leads nicely to next section).

@@

\subsection{The right statistics}
\label{ssec:rightstatistics}

In this section, I briefly motivate the choice of statistical models which I use later in Section~\ref{sec:corpusmodels}.
In the spirit of the careful and methodologically sound approach to cognitively oriented corpus linguistics laid out in Section~\ref{sec:cogocl}, authors should always provide at least a minimal motivation as part of the standard protocol, especially if there are alternative methods.

% NO to exploratory methods if we have hypotheses
% Testing (hypotheses about signs of regressors)
% Estimation => bootstrapping
% Model selection
%  => random vs fixed is not so simple, rather matter of getting model to converge
%  => Maximality
%  => NO to Gries 2015
% Model evaluation
% VS: Bayesian hype (Levshina)
% VS: alternative math (R-W)

@@

%In the following, I examine the case alternation in German measure NPs under a cognitively oriented corpus linguistic perspective.
%Since, however, the statistical tool kit (hierarchical generalised linear models, general tools such as $p$ values and confidence intervals, resampling techniques, model selection and model evaluation strategies, alternative algorithms for estimation, etc.) is heterogeneous and often criticised (especially under noisy conditions and when researcher run the risk of doing data mining instead of proper theory testing) I present a careful protocol for specifying, estimating, validating, and interpreting such models in Section \ref{sec:corpusmodels}.

Before discussing the results in Section~\ref{sec:corpusmodels}, I show how I made sure that the relevant data is contained in the sample in Section \ref{sec:gettingdata}.

@@

%\subsection{Tools for modelling alternations}
%\label{sec:righttools}
%
%In this section, I discuss the relevant statistical approaches to modelling alternations based on corpus data.
%I mainly focus on well-established Generalised Linear Models (GLMs) and their \textit{multi-level} or \textit{hierarchical} extensions, which are also called \textit{mixed models} (GLMMs).
%
%% TODO CONTENT HERE
%
%In Section~\ref{ssec:corpusnonhierarchicalmodel}, I briefly discuss why a non-hierarchical model -- while it is an option in principle -- has serious disadvantages for the data set under discussion (and similar data sets).
%In Section~\ref{ssec:corpushierarchicalmodel}, I present the hierarchical model which is actually interpreted as the primary result of this study.
%Section~\ref{ssec:bayesian} takes a short detour and shows (by example) that results from Bayesian estimation are not intrinsically different from those obtained from Maximum Likelihood methods, although Bayesian approaches have recently been touted to solve various problems facing applied statistics in various fields, even including cognitively oriented corpus linguistics.
%Finally, I summarise and interpret the results of the corpus study in Section~\ref{ssec:modelssummary}.
